---
title: "Introduction to spharmlithic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to spharmlithic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

**spharmlithic** supports two complementary lines of quantitative analysis on 3D stone artefacts:

- **Morphological analysis.** Spherical harmonic decomposition of 3D mesh surfaces (.stl), capturing overall artefact shape.
- **Scar pattern analysis.** Alignment, descriptive statistics, and spherical harmonic decomposition of scar orientation vectors, capturing how flaking removals are organised during core reduction.

This vignette is a **getting-started guide**: it walks through both workflows using the small example data bundled with the package, taking you from raw artefacts to spherical harmonic coefficients and an exported results table. Once you have that table, the companion vignette `vignette("Statistics", package = "spharmlithic")` shows how to analyse it statistically — exploration with PCA, group comparison with PERMANOVA, and tests of standardisation.

------------------------------------------------------------------------

## 1. Setup

### Installing from R

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("PeiyuanXiao/spharmlithic")
```

### Installing the Python back-end

Spherical harmonic analysis requires a Python environment with `pyshtools` and related packages. If you only need alignment and descriptive statistics you can skip this step.

```{r install-python, eval = FALSE}
library(spharmlithic)

# Core install (scar pattern analysis)
install_spharmlithic_python()

# Full install (also enables morphological analysis from meshes)
install_spharmlithic_python(mesh = TRUE)
```

### Activating the Python environment

At the start of each R session, point `reticulate` to the conda environment:

```{r load-package, eval = TRUE, message = FALSE}
library(spharmlithic)
library(readxl)
```

```{r activate-python}
use_spharmlithic_python("r-spharmlithic")
```

> **First-time users:** if you have not yet created the conda environment, run `install_spharmlithic_python()` first. See `?install_spharmlithic_python` for details.

### Docker (recommended for macOS)

On **macOS** (especially Apple Silicon), `spharm_from_meshes()` may crash the R session because `open3d`'s OpenGL back-end is incompatible. A pre-built Docker image avoids this entirely:

```{bash docker, eval = FALSE}
docker pull peiyuanxiao/spharmlithic
docker run -d -p 8787:8787 \
  -v /path/to/your/data:/home/rstudio/data \
  peiyuanxiao/spharmlithic
```

Open <http://localhost:8787> in your browser (user: `rstudio`, password: `spharm`). The container includes R, RStudio Server, and the full Python environment pre-configured. Example data and a quick-start script are in `~/examples/`.

`spharm_from_directions()` works on all platforms without Docker.

------------------------------------------------------------------------

## 2. Input Data Format

### Scar orientation data (Excel / CSV)

The scar analysis pipeline expects a data frame with **one row per scar**, where all scars from all specimens are stacked together.

**Required columns:**

| Column                          | Description                         |
|:--------------------------------|:------------------------------------|
| `ID`                            | Specimen identifier                 |
| `Start_X`, `Start_Y`, `Start_Z` | XYZ coordinates of scar start-point |
| `End_X`, `End_Y`, `End_Z`       | XYZ coordinates of scar end-point   |

**Additional columns required by `align_morph_batch()` (Lin 2024 pipeline):**

| Column | Description |
|:-----------------------------------|:-----------------------------------|
| `Norm_X`, `Norm_Y`, `Norm_Z` | Morphological best-fitted plane normal vector |
| `Pos_X`, `Pos_Y`, `Pos_Z` | Centroid of the morphological best-fitted plane |

**Optional columns:**

| Column | Description |
|:-----------------------------------|:-----------------------------------|
| `Scar_ID` | Individual scar identifier within a specimen |
| `Length` | Pre-computed scar length (if absent, computed from start/end coordinates) |
| `Typology` | Specimen type label (e.g. "Levallois", "Discoid"); used for grouping in the interactive SPHARM viewer |

**Minimal example (3 scars from 2 specimens):**

| ID   | Start_X | Start_Y | Start_Z | End_X | End_Y | End_Z |  Typology |
|:-----|--------:|--------:|--------:|------:|------:|------:|----------:|
| S001 |    10.2 |     5.1 |     3.0 |  12.4 |   6.3 |   2.8 | Levallois |
| S001 |    11.0 |     4.8 |     3.1 |  13.5 |   5.9 |   2.5 | Levallois |
| S002 |    20.1 |     8.3 |     1.5 |  22.0 |   9.1 |   1.2 |   Discoid |

The bundled example data follows this format:

```{r example-data, eval = TRUE}
scar_path <- system.file("extdata", "example_scars.xlsx",
                          package = "spharmlithic")
example   <- readxl::read_excel(scar_path)
names(example)
```

### 3D mesh files

Place `.stl` mesh files in a single directory. Each file is treated as one specimen; the filename (without extension) becomes the specimen ID.

**Mesh preparation recommendations:**

- Simplify meshes to 50,000–200,000 faces in advance using dedicated 3D software (e.g. MeshLab, CloudCompare). The package includes automatic pre-decimation for very high-resolution models (\>3M faces), but pre-simplification significantly reduces computation time.
- Meshes must be **watertight** (closed, manifold, genus-0). Repair any holes, non-manifold edges, or self-intersections before analysis.

```{r mesh-dir, eval = TRUE}
stl_dir <- system.file("extdata", "meshes", package = "spharmlithic")
list.files(stl_dir, pattern = "\\.stl$", ignore.case = TRUE)
```

### Output columns after alignment

Both alignment pipelines (`align_scar_batch()` and `align_morph_batch()`) add nine columns to the input data frame:

| Column | Description |
|:-----------------------------------|:-----------------------------------|
| `s_x`, `s_y`, `s_z` | Aligned scar start-point coordinates |
| `e_x`, `e_y`, `e_z` | Aligned scar end-point coordinates |
| `d_x`, `d_y`, `d_z` | Aligned unit direction vectors — the primary input for `compute_SPI()`, `compute_EI()`, and `spharm_from_directions()` |

------------------------------------------------------------------------

## 3. Scar Pattern Analysis

### 3.1 Load the example data

```{r load-scar-data, eval = TRUE}
scar_path <- system.file("extdata", "example_scars.xlsx",
                          package = "spharmlithic")
raw_data  <- read_excel(scar_path)
head(raw_data)
```

### 3.2 Alignment

Before comparing scar patterns across specimens, the orientation vectors need to be rotated into a common coordinate frame. **spharmlithic** provides two alignment pipelines.

#### SVD alignment (`align_scar_batch`)

The SVD method defines three principal axes from the full set of scar direction vectors via singular value decomposition. It is general-purpose and does not require a predefined morphological axis.

```{r align-svd, eval = TRUE}
aligned_svd <- align_scar_batch(raw_data)
head(aligned_svd)
```

#### Lin 2024 alignment (`align_morph_batch`)

The method of Lin et al. (2024) first orients the specimen along a morphological long-axis and then rotates around that axis to align the dominant flaking direction. It requires `Norm_X/Y/Z` and `Pos_X/Y/Z` columns in the input data.

```{r align-lin, eval = TRUE}
aligned_lin <- align_morph_batch(raw_data)
head(aligned_lin)
```

Both pipelines produce output in the same format, so all downstream functions work identically regardless of which alignment you choose. We use the SVD-aligned data for the rest of this vignette.

### 3.3 Descriptive statistics

#### Scar Pattern Index (SPI)

The SPI quantifies the overall directionality of flaking removals. Higher values indicate a more unidirectional pattern.

```{r spi, eval = TRUE}
spi <- compute_SPI(aligned_svd$d_x, aligned_svd$d_y, aligned_svd$d_z)
spi
```

An angular variant is also available:

```{r spi-angle, eval = TRUE}
spi_angle <- compute_spi_angle(aligned_svd$d_x, aligned_svd$d_y, aligned_svd$d_z)
spi_angle
```

For the original length-weighted SPI (Clarkson et al. 2006), pass scar lengths:

```{r spi-weighted, eval = TRUE}
lens <- get_scar_length(aligned_svd)
spi_w <- compute_SPI(aligned_svd$d_x, aligned_svd$d_y, aligned_svd$d_z,
                     lengths = lens)
spi_w
```

#### Elongation and Isotropy

`compute_EI()` returns the elongation (E) and isotropy (I) ratios derived from the orientation tensor eigenvalues (Lin et al. 2024):

```{r ei, eval = TRUE}
ei <- compute_EI(aligned_svd$d_x, aligned_svd$d_y, aligned_svd$d_z)
ei
```

#### Per-specimen statistics

Use `dplyr::group_by()` to compute statistics for each specimen:

```{r per-specimen, eval = FALSE}
library(dplyr)
aligned_svd %>%
  group_by(ID) %>%
  summarise(
    SPI       = compute_SPI(d_x, d_y, d_z),
    SPI_angle = compute_spi_angle(d_x, d_y, d_z),
    compute_EI(d_x, d_y, d_z)
  )
```

### 3.4 Alignment visualisation

To export an interactive HTML report showing the alignment steps:

```{r export-html-svd}
export_alignment_html_svd(raw_data, out_path = "alignment_svd.html")
export_alignment_html_lin2024(raw_data, out_path = "alignment_lin2024.html")
```

### 3.5 Spherical harmonic analysis

With the Python back-end active, scar orientation vectors can be decomposed into spherical harmonic coefficients. This applies von Mises–Fisher kernel density estimation on the sphere and then fits the harmonic expansion:

```{r spharm-scar}
sh_scar <- spharm_from_directions(aligned_svd, lmax = 20, bandwidth = 0.35)
sh_scar
```

To convert the results into a flat data frame for downstream multivariate analysis (PCA, clustering, distance matrices):

```{r df-scar}
df_scar <- spharm_to_dataframe(sh_scar)
head(df_scar)
```

------------------------------------------------------------------------

## 4. Morphological Analysis

Morphological analysis takes 3D mesh files (.stl) as input and characterises overall artefact shape via spherical harmonics. This requires the mesh extension: `install_spharmlithic_python(mesh = TRUE)`.

> **macOS users:** `spharm_from_meshes()` may crash the R session on macOS due to `open3d` OpenGL incompatibilities. Use the Docker image instead (see Section 1).

```{r spharm-morph}
stl_dir  <- system.file("extdata", "meshes", package = "spharmlithic")
sh_morph <- spharm_from_meshes(stl_dir, lmax = 20)
sh_morph
```

To convert to a flat data frame:

```{r df-morph}
df_morph <- spharm_to_dataframe(sh_morph)
head(df_morph)
```

------------------------------------------------------------------------

## 5. Interactive SPHARM Viewer

`export_spharm_html()` generates a self-contained HTML file with an interactive Three.js-based 3D viewer for exploring spherical harmonic reconstructions. The viewer synthesises the harmonic expansion in real time on the client side — only the coefficients are embedded, so file sizes stay small.

You can pass results from either analysis, or both:

```{r viewer-scar-only}
# Scar pattern only
export_spharm_html(
  scar     = sh_scar,
  out_path = "viewer_scar.html"
)
```

```{r viewer-both}
# Both morphology and scar pattern side-by-side
export_spharm_html(
  morph    = sh_morph,
  scar     = sh_scar,
  out_path = "spharm_viewer.html"
)
```

### Specimen metadata and grouping

An optional `meta` argument accepts a data frame with specimen metadata. If a `Typology` column is present, the viewer groups the specimen dropdown by type and enables a Type Mean toggle that averages coefficients within each group:

```{r viewer-meta}
meta <- data.frame(
  ID       = c("ClarksonEXP01_Levallois", "ClarksonEXP02_Discoid"),
  Typology = c("Levallois", "Discoid")
)
export_spharm_html(
  scar     = sh_scar,
  meta     = meta,
  out_path = "spharm_viewer.html",
  title    = "Example specimens"
)
```

### Viewer features

- **Dual viewports** — morphology and scar direction side-by-side, with synchronised rotation.
- **Degree slider and animation** — l = 1 → lmax, for visualising how successive harmonic degrees build up the reconstruction.
- **Six material presets** — Ceramic, Clay, Glass, Brushed Metal, X-Ray, Flat — plus a radial deviation colormap.
- **Type Mean** — average coefficients across specimens sharing the same Typology.
- **PNG screenshot** and **OBJ mesh export** at the current degree.

------------------------------------------------------------------------

## 6. Harmonic Degree (`lmax`)

The maximum spherical harmonic degree controls the level of detail in the decomposition. The default is `lmax = 20`, which captures the dominant features for most artefacts.

**Morphological analysis** can benefit from higher degrees if you need finer geometric detail:

```{r lmax-morph}
sh_fine <- spharm_from_meshes(stl_dir, lmax = 50)
```

Note that computation time and output size grow substantially with higher degrees (the number of coefficients scales as `2 × (lmax + 1)²`). When increasing `lmax` for morphology, you may also want to increase `grid_size` (default 256) for adequate sampling.

**Scar pattern analysis** does not benefit from degrees above 20. The underlying von Mises–Fisher KDE already smooths the directional density, so increasing `lmax` beyond 20 adds noise fitting rather than meaningful detail. If you do increase `lmax`, the DH grid size must satisfy `dh_size >= 2 * (lmax + 1)`.

------------------------------------------------------------------------

## 7. Low-level Reconstruction

For users who need the reconstructed density as numerical data rather than an interactive visualisation, `spharm_reconstruct()` performs the inverse spherical harmonic transform and returns:

- `density` — numeric matrix (grid_size × 2 × grid_size)
- `lon`, `lat` — longitude and latitude vectors (radians)
- `xyz` — unit-sphere Cartesian coordinates

```{r reconstruct}
recon <- spharm_reconstruct(
  sh_scar[["ClarksonEXP01_Levallois"]]$coefficients,
  grid_size = 64
)
```

This is useful for computing point-wise differences between specimens, extracting density peaks, or building custom visualisations:

```{r plotly-recon}
library(plotly)
nlat <- length(recon$lat)
nlon <- length(recon$lon)
x_mat <- matrix(recon$xyz[, "x"], nlat, nlon)
y_mat <- matrix(recon$xyz[, "y"], nlat, nlon)
z_mat <- matrix(recon$xyz[, "z"], nlat, nlon)

plot_ly(x = x_mat, y = y_mat, z = z_mat,
        surfacecolor = recon$density,
        type = "surface", colorscale = "Hot",
        showscale = TRUE)
```

------------------------------------------------------------------------

## References

Bretzke, K., & Conard, N. J. (2012). Evaluating morphological variability in lithic assemblages using 3D models of stone artifacts. *Journal of Archaeological Science*, 39(12), 3741–3749.

Clarkson, C., Vinicius, L., & Lahr, M. M. (2006). Quantifying flake scar patterning on cores using 3D recording techniques. *Journal of Archaeological Science*, 33(1), 132–142.

Lin, S. C., Clarkson, C., Julianto, I. M. A., Ferdianto, A., & Sutikna, T. (2024). A new method for quantifying flake scar organisation on cores using orientation statistics. *Journal of Archaeological Science*, 167, 105998.

McPherron, S. P. (2018). Additional statistical and graphical methods for analyzing site formation processes using artifact orientations. *PLoS ONE*, 13(1), e0190195.

Wieczorek, M. A., & Meschede, M. (2018). SHTools: Tools for working with spherical harmonics. *Geochemistry, Geophysics, Geosystems*, 19(8), 2574–2592.

Ye, Z., Pei, S. W., Ma. D. D., Li, H., Marwick, B. (2026). Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China). *Journal of Archaeological Science*, 190, 106551.
