---
title: "Introduction to spharmlithic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to spharmlithic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

**spharmlithic** supports two complementary lines of quantitative
analysis on 3D stone artefacts:

- **Track A -- 3D morphology.** Spherical harmonic decomposition of
  3D mesh surfaces (.stl/.ply), capturing overall artefact shape.
- **Track B -- Scar patterns.** Alignment, descriptive statistics,
  and spherical harmonic decomposition of scar orientation vectors,
  capturing how flaking removals are organised on core reduction.

This vignette walks through both workflows using the small example data
bundled with the package.

## Setup

```{r load-package, eval = TRUE, message = FALSE}
library(spharmlithic)
library(readxl)
```

If you need spherical harmonic analysis (either track), activate the
Python back-end once per session:

```{r activate-python}
use_spharmlithic_python("r-spharmlithic")
```

> **First-time users:** if you have not yet created the conda
> environment, run `install_spharmlithic_python()` (and optionally
> `install_spharmlithic_python(mesh = TRUE)` for Track A). See
> `?install_spharmlithic_python` for details.

------------------------------------------------------------------------

## Track B -- Scar pattern analysis

### 1. Load the example data

The package ships with a small Excel file containing scar orientation
data for two experimental specimens (a Levallois core and a Discoid
core):

```{r load-scar-data, eval = TRUE}
scar_path <- system.file("extdata", "example_scars.xlsx",
                          package = "spharmlithic")
raw_data  <- read_excel(scar_path)
head(raw_data)
```

### 2. Alignment

Before comparing scar patterns across specimens, the orientation vectors
need to be rotated into a common coordinate frame. **spharmlithic**
provides two alignment pipelines.

#### SVD alignment

The SVD method defines three principal axes from the full set of scar
direction vectors via singular value decomposition. It is
general-purpose and does not require a predefined morphological axis.

```{r align-svd, eval = TRUE}
aligned_svd <- align_scar_batch(raw_data)
head(aligned_svd)
```

#### Lin 2024 alignment

The method of Lin et al. (2024) first orients the specimen along a
morphological long-axis and then rotates around that axis to align the
dominant flaking direction. It is best suited when a meaningful
technological axis (e.g. the platform-base axis) can be identified.

```{r align-lin, eval = TRUE}
aligned_lin <- align_morph_batch(raw_data)
head(aligned_lin)
```

Both pipelines produce output in the same format, so all downstream
functions work identically regardless of which alignment you choose. We
use the SVD-aligned data for the rest of this vignette.

### 3. Descriptive statistics

#### Scar Pattern Index (SPI)

The SPI quantifies the overall directionality of flaking removals.
Higher values indicate a more unidirectional pattern.

```{r spi, eval = TRUE}
spi <- compute_SPI(aligned_svd$d_x, aligned_svd$d_y, aligned_svd$d_z)
spi
```

An angular variant is also available:

```{r spi-angle, eval = TRUE}
spi_angle <- compute_spi_angle(aligned_svd$d_x, aligned_svd$d_y, aligned_svd$d_z)
spi_angle
```

#### Elongation and Isotropy

The EI captures the elongation ratio and isotropy ratio of the scar
pattern:

```{r ei, eval = TRUE}
ei <- compute_EI(aligned_svd$d_x, aligned_svd$d_y, aligned_svd$d_z)
ei
```

### 4. Visualisation

Interactive 3D plots can be generated with the built-in Plotly helpers:

```{r vis-3d}
# Build a multi-panel figure for the SVD alignment
panels <- build_panel_scar(demo_id = "ClarksonEXP01", raw_data = raw_data)
panels
```

To export an interactive HTML report:

```{r export-html}
export_alignment_html_svd(aligned_svd, file = "alignment_report.html")
```

### 5. Spherical harmonic analysis (Track B)

With the Python back-end active, scar orientation vectors can be
decomposed into spherical harmonic coefficients. This applies von
Mises-Fisher kernel density estimation on the sphere and then fits
the harmonic expansion:

```{r spharm-trackb}
sh_scar <- spharm_from_directions(aligned_svd, lmax = 20, bandwidth = 0.35)
sh_scar
```

The density surface can be reconstructed on a regular grid:

```{r reconstruct-trackb}
recon <- spharm_reconstruct(sh_scar)
recon
```

To convert the results into a flat data frame for further analysis in R:

```{r df-trackb}
df <- spharm_to_dataframe(sh_scar)
head(df)
```

------------------------------------------------------------------------

## Track A -- Morphological analysis

Track A takes 3D mesh files (.stl/.ply) as input and characterises
overall artefact shape via spherical harmonics. This requires the mesh
extras (`install_spharmlithic_python(mesh = TRUE)`).

### 1. Locate the example meshes

```{r mesh-dir, eval = TRUE}
stl_dir <- system.file("extdata", "meshes", package = "spharmlithic")
list.files(stl_dir, pattern = "\\.(stl|ply)$")
```

### 2. Spherical harmonic analysis (Track A)

`spharm_from_meshes()` reads each mesh file, extracts surface normals,
maps them onto the unit sphere, and computes spherical harmonic
coefficients:

```{r spharm-tracka}
sh_morph <- spharm_from_meshes(stl_dir, lmax = 20)
sh_morph
```

The same downstream functions apply:

```{r downstream-tracka}
recon_morph <- spharm_reconstruct(sh_morph)
df_morph    <- spharm_to_dataframe(sh_morph)
head(df_morph)
```

------------------------------------------------------------------------

## Comparing the two tracks

| | Track A -- 3D morphology | Track B -- Scar patterns |
|:---|:---|:---|
| **Input** | 3D mesh files (.stl/.ply) | Scar orientation vectors (Excel / data frame) |
| **What it captures** | Overall artefact shape | Organisation of flaking removals |
| **Key function** | `spharm_from_meshes()` | `spharm_from_directions()` |
| **Python extras** | Requires `mesh = TRUE` | Base Python install |
| **Typical use** | Shape-based typology, reduction intensity | Flaking strategy comparison |

Both tracks produce spherical harmonic coefficient objects that can be
reconstructed (`spharm_reconstruct()`) and converted to data frames
(`spharm_to_dataframe()`), so they integrate into the same analytical
pipeline.

------------------------------------------------------------------------

## Using your own data

### Scar orientation data (Track B)

Prepare an Excel or CSV file with the same column structure as the
example data. At minimum you need columns identifying the specimen, the
scar, and the three direction-vector components (d_x, d_y, d_z). See
`?align_scar_batch` for the expected format.

### 3D meshes (Track A)

Place your mesh files in a single directory and pass the path to
`spharm_from_meshes()`:

```{r own-data}
sh_own <- spharm_from_meshes("path/to/my/mesh_files", lmax = 20)
```

------------------------------------------------------------------------

## References

Bretzke, K., & Conard, N. J. (2012). Evaluating morphological
variability in lithic assemblages using 3D models of stone artifacts.
*Journal of Archaeological Science*, 39(12), 3741-3749.

Clarkson, C., Vinicius, L., & Lahr, M. M. (2006). Quantifying flake
scar patterning on cores using 3D recording techniques. *Journal of
Archaeological Science*, 33(1), 132-142.

Lin, S. C., Clarkson, C., Julianto, I. M. A., Ferdianto, A., &
Sutikna, T. (2024). A new method for quantifying flake scar
organisation on cores using orientation statistics. *Journal of
Archaeological Science*, 167, 105998.

McPherron, S. P. (2018). Additional statistical and graphical methods
for analyzing site formation processes using artifact orientations.
*PLoS ONE*, 13(1), e0190195.

Wieczorek, M. A., & Meschede, M. (2018). SHTools: Tools for working
with spherical harmonics. *Geochemistry, Geophysics, Geosystems*,
19(8), 2574-2592.

Ye, Z., Pei, S. W., Ma. D. D., Li, H., Marwick, B. (2026). Spherical
harmonic analysis of faceted spheroids identifies shaping strategies
and standardisation at Qianshangying (North China). *Journal of
Archaeological Science*, 190, 106551.
