---
title: "Statistical Analysis of Spherical Harmonic Data"
author: "Peiyuan Xiao, Ben Marwick"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Statistical analysis of spherical harmonic data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# This vignette uses several Suggested packages. If any are missing we still
# render the prose and show the code, but skip evaluation, so the vignette
# never fails to build.
can_run <- requireNamespace("vegan",        quietly = TRUE) &&
           requireNamespace("compositions", quietly = TRUE) &&
           requireNamespace("ggplot2",      quietly = TRUE) &&
           requireNamespace("tidyr",        quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 6,
  fig.height = 4,
  eval     = can_run
)
```

## Introduction

The goal of this vignette is to walk you through using the spharmlithic package to tackle three research questions that are central to many analyses of stone artefact assemblages from archaeological sites:

1.  **Exploration**: how is artefact shape/scar pattern variation structured across specimens? We will demonstrate how to use Principal Component Analysis (PCA) for this.
2.  **Group difference**: do artefacts made by different reduction strategies differ in their SPHARM descriptors? PERMANOVA is a classic statistical method for this question. 
3.  **Standardisation**: is one reduction strategy more standardised (less variable) than another? Multivariate dispersion (PERMIDSP) is useful to answer this question. 

This vignette follows from our companion vignette `vignette("Introduction", package = "spharmlithic")` that walks through how to turn 3D artefacts into spherical harmonic (SH) coefficients with `spharm_from_directions()` and `spharm_from_meshes()`, and how to flatten the result with `spharm_to_dataframe()`. Here we demonstrate functions in the package that are designed for typical research applications. 

Our choice of statistical methods for this walk-through is informed by three features of SPHARM power spectra:

- Power spectra are rotation-invariant. Each `power_lN` value summarises the energy at spherical harmonic degree `N`, independent of how the specimen was oriented. This makes a power spectrum a robust feature set for cross-specimen comparison, so we use it throughout this workflow.
- Power spectra are compositional data. Each specimen's spectrum is closed to a constant sum (here, 1): the values are relative proportions of energy across degrees, not absolute magnitudes. Compositional data live on the simplex, not in Euclidean space, so applying Euclidean-based methods (PCA, distances) to the raw proportions induces spurious negative correlations (the closure problem). To fix this we use the isometric log-ratio (ILR) transform (Egozcue et al. 2003), which maps the composition into ordinary Euclidean coordinates where standard methods are valid. Distances computed on ILR coordinates are Aitchison distances.
- The number of variables measured on each artefact via spherical harmonics power spectra, `p`,  is often comparable to or larger than `n`. A typical power spectrum may have 20+ columns. Classical MANOVA becomes unstable and unreliable when the number of variables approaches the number of specimens. To avoid this we use distance- and permutation-based methods (PERMANOVA, `betadisper`) and dimension reduction (PCA).

## The Example Data

The spharmlithic package bundles a worked example: spherical harmonic power spectra for 58 experimental cores knapped in five known reduction strategies by Professor Chris Clarkson. These data are included here with his permission. In this section we load the data, two parallel tables that describe the artefacts in two domains: scar direction (how flaking removals are organised) and morphology (overall shape). These tables were produced by following the steps in our Introduction vignette (`spharm_from_*()` → `spharm_to_dataframe()`). We also include with the package the flattened tables resulting from the Introduction vignette. This means that all the code in this statistical analysis vignette can be run without needing the Python backend. 

```{r load-data}
# load the tables
scar  <- read.csv(
  system.file("extdata", "exp_cores_scar.csv",  package = "spharmlithic"),
  check.names = FALSE)
morph <- read.csv(
  system.file("extdata", "exp_cores_morph.csv", package = "spharmlithic"),
  check.names = FALSE)

# set metadata
strategy_levels <- c("Unidirectional", "Bidirectional",
                     "Levallois", "Discoid", "Multiplatform")
scar$Typology  <- factor(scar$Typology,  levels = strategy_levels)
morph$Typology <- factor(morph$Typology, levels = strategy_levels)

# prepare a palette for visualization
typology_colors <- c(
  "Unidirectional" = "#BA8530",
  "Bidirectional"  = "#788C4A",
  "Levallois"      = "#4A6E8A",
  "Discoid"        = "#802520",
  "Multiplatform"  = "#8A7A68"
)

# display summary table
knitr::kable(
  as.data.frame(table(Strategy = scar$Typology)),
  col.names = c("Reduction strategy", "n"),
  caption   = "Experimental cores per reduction strategy")
```

Each of the two tables has an `ID`, a `Typology`, and the power spectrum from degree 1 to 20 (`power_l1` … `power_l20`; degree 0, the constant term, is dropped because it carries no shape information). Each row sums to 1:

```{r preview}
knitr::kable(
  head(scar[, 1:6]),
  caption = "First few rows of the scar-direction power spectra")

# Confirm the spectra are closed (compositional)
range(rowSums(scar[, grep("^power_l", names(scar))]))
```

To prepare for answering our research questions in the next sections, we separate the numeric feature matrix from the grouping variable:

```{r feature-matrices}
scar_cols   <- paste0("power_l", 1:6) # first 6 degrees of the power spectrum
morph_cols  <- paste0("power_l", 1:8) # first 8 degrees of the power spectrum
scar_feat   <- scar[,  scar_cols]
morph_feat  <- morph[, morph_cols]
groups      <- scar$Typology     
```

## Compositional Data Analyses of the Power Spectra

Before any Euclidean-based analysis, we move the spectra off the simplex with the ILR transform. ILR (like any log-ratio) requires strictly positive values, but the scar spectra contain exact zeros (degrees with no energy). We handle them with the function `make_ilr` which includes a multiplicative zero replacement function that preserves the unit sum `replace_zeros`, following Xiao et al (in preparation):

```{r ilr}
# ILR coordinates: D parts -> (D - 1) real-valued coordinates per specimen.
scar_ilr  <- make_ilr(scar_feat)    # 6 degrees -> 5 ILR coordinates
morph_ilr <- make_ilr(morph_feat)   # 8 degrees -> 7 ILR coordinates

# first few rows
head(scar_ilr)
```

The output is a data frame with one row per artefact, and one column for each ILR coordinate (the transformed power spectra). All of the analyses below operate on these ILR coordinates. Note that this is specific to the power spectrum (a composition); the **raw coefficients** returned by `spharm_to_dataframe(include_coeffs = TRUE)` are ordinary real-valued numbers and would be analysed directly, not log-ratio transformed.

## How to choose an appropriate number of degrees from the power spectra?

The feature matrices defined above keep only the first 6 degrees for scar direction and the first 8 for morphology, rather than all 20. This section shows the diagnostic behind that choice. 

Although coefficients were computed to degree 20 for faithful reconstruction, retaining all 20 degrees for statistical comparison would add dimensionality and, for scar patterns, noise. We therefore screen degrees prior to all statistical analyses using two diagnostics: the across-specimen coefficient of variation (CV), indexing the reliability of each degree's between-specimen signal, and cumulative power, indexing the share of total variance each degree carries.

We provide the function `degree_diagnostics` to operate on the full 20-degree spectra and calculate per-degree coefficients of variation and cumulative mean power from a power spectrum before the truncation used in the next steps.

```{r degree-selection, fig.width = 7, fig.height = 3.4, fig.cap = "Per-degree diagnostics for the two power spectra (experimental cores, n = 58): across-specimen coefficient of variation (left) and cumulative power (right), each overlaying the two domains. Red dashed lines mark the retained truncations (degree 6 for scar direction, degree 8 for morphology); grey dashed lines mark CV = 100% and the 95% / 99% cumulative-power levels."}
library(ggplot2)

deg_df <- rbind(
  degree_diagnostics(scar,  "SP-SPHARM (scar direction)"),
  degree_diagnostics(morph, "M-SPHARM (morphology)"))
deg_df$descriptor <- factor(deg_df$descriptor,
  levels = c("M-SPHARM (morphology)", "SP-SPHARM (scar direction)"))

descriptor_colors <- c(
  "M-SPHARM (morphology)"      = "#4A6E8A",
  "SP-SPHARM (scar direction)" = "#BA8530")

# Long form so the two diagnostics share one faceted figure.
deg_long <- tidyr::pivot_longer(deg_df, c(cv_pct, cumul_pct),
                                names_to = "metric", values_to = "value")
deg_long$metric <- factor(deg_long$metric, c("cv_pct", "cumul_pct"),
  labels = c("Across-specimen CV (%)", "Cumulative power (%)"))

# Panel-specific reference lines: CV = 100% (noise threshold), and the 95 / 99%
# cumulative-power levels.
ref_lines <- data.frame(
  metric = factor(c("Across-specimen CV (%)",
                    "Cumulative power (%)", "Cumulative power (%)"),
                  levels = levels(deg_long$metric)),
  yint   = c(100, 95, 99))

ggplot(deg_long, aes(degree, value, colour = descriptor)) +
  geom_hline(data = ref_lines, aes(yintercept = yint),
             linetype = "dashed", colour = "grey55", linewidth = 0.3) +
  geom_vline(xintercept = c(6, 8), colour = "red",
             linetype = "dashed", linewidth = 0.3) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.4) +
  scale_colour_manual(values = descriptor_colors) +
  scale_x_continuous(breaks = seq(2, 20, 2)) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(x = "Spherical harmonic degree", y = NULL, colour = NULL) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom")
```

The two spectra behave differently and are truncated on different grounds. For scar direction (SP-SPHARM), power is effectively exhausted by the sixth degree (cumulative energy ≈ 99.9%), and higher degrees are noise-dominated (CV \> 100% from degree 9 onward); we therefore retain degrees 1–6. For morphology (M-SPHARM), between-specimen signal remains reliable across all degrees (CV \< 90% throughout), but variance is strongly concentrated in the low degrees, with \~98% of cumulative power captured by the eighth degree; we retain degrees 1–8, since the reliable but very low-energy higher degrees (together under 2% of power) would inflate dimensionality and dilute multivariate comparisons without materially improving discrimination.

## 1. Exploring variation with PCA

A quick way to see whether the strategies separate at all is to project each domain's ILR coordinates onto their first two principal components. We run PCA on the **unscaled** coordinates (no `scale. = TRUE`): ILR already lays the composition out in a Euclidean space whose distances are the Aitchison distances the PERMANOVA below relies on, so standardising each coordinate would reweight the parts and distort that geometry.

```{r pca}
pca_scar  <- prcomp(scar_ilr)
pca_morph <- prcomp(morph_ilr)

# Variance explained by the first few components, both domains
var_tab <- data.frame(
  PC = paste0("PC", 1:5),
  `Scar direction (%)` = round(100 * summary(pca_scar)$importance[2, 1:5], 1),
  `Morphology (%)`     = round(100 * summary(pca_morph)$importance[2, 1:5], 1),
  check.names = FALSE)
knitr::kable(var_tab, row.names = FALSE,
  caption = "Variance explained by the leading principal components, by domain")
```

```{r pca-plot, fig.cap = "Cores in the space of the first two principal components of the ILR coordinates, by domain, coloured by reduction strategy, with convex hulls outlining each group. Axis scales are free; variance explained is in the table above."}
library(ggplot2)

pca_df <- rbind(
  data.frame(Domain = "Scar direction",
             PC1 = pca_scar$x[, 1],  PC2 = pca_scar$x[, 2],  Strategy = groups),
  data.frame(Domain = "Morphology",
             PC1 = pca_morph$x[, 1], PC2 = pca_morph$x[, 2], Strategy = groups))
pca_df$Domain   <- factor(pca_df$Domain, levels = c("Scar direction", "Morphology"))
pca_df$Strategy <- factor(pca_df$Strategy, levels = strategy_levels)

# Convex hulls per strategy, matching the project's LDA / ternary scatterplots
hull_df <- do.call(rbind, lapply(
  split(pca_df, list(pca_df$Domain, pca_df$Strategy), drop = TRUE),
  function(d) d[chull(d$PC1, d$PC2), ]))

ggplot(pca_df, aes(PC1, PC2, colour = Strategy)) +
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.35, linetype = "dashed") +
  geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.35, linetype = "dashed") +
  geom_polygon(data = hull_df, aes(fill = Strategy, group = Strategy),
               alpha = 0.25, colour = NA) +
  geom_point(size = 2, alpha = 0.85, shape = 16) +
  scale_colour_manual(values = typology_colors) +
  scale_fill_manual(values  = typology_colors) +
  facet_wrap(~ Domain, scales = "free") +
  labs(x = "PC1", y = "PC2") +
  theme_bw() +
  theme(panel.grid = element_blank())
```

The convex hulls give a first impression of both **location** (do the groups sit in different places?) and **spread** (are some groups tighter than others?), the two things we test formally below.

It is also informative to look at the **mean power spectrum** of each strategy (on the original closed scale): the degree at which energy concentrates reflects the angular scale of the dominant structure.

```{r spectrum-plot, fig.cap = "Mean power spectrum by reduction strategy and domain (ribbon = +/- 1 SE). Values are closed proportions."}
library(tidyr)

# The two domains now keep different numbers of degrees, so pivot each to long
# form (a shared schema) before combining.
spec_long <- rbind(
  data.frame(Domain = "Scar direction", Strategy = groups, scar_feat,  check.names = FALSE) |>
    tidyr::pivot_longer(cols = all_of(scar_cols),  names_to = "degree", values_to = "power"),
  data.frame(Domain = "Morphology",     Strategy = groups, morph_feat, check.names = FALSE) |>
    tidyr::pivot_longer(cols = all_of(morph_cols), names_to = "degree", values_to = "power"))
spec_long$degree   <- as.integer(sub("power_l", "", spec_long$degree))
spec_long$Domain   <- factor(spec_long$Domain, levels = c("Scar direction", "Morphology"))
spec_long$Strategy <- factor(spec_long$Strategy, levels = strategy_levels)

spec_summ <- aggregate(power ~ Domain + Strategy + degree, spec_long,
                       FUN = function(x) c(m = mean(x),
                                           se = sd(x) / sqrt(length(x))))
spec_summ <- do.call(data.frame, spec_summ)

ggplot(spec_summ, aes(degree, power.m, colour = Strategy, fill = Strategy)) +
  geom_ribbon(aes(ymin = power.m - power.se, ymax = power.m + power.se),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.6) +
  scale_colour_manual(values = typology_colors) +
  scale_fill_manual(values  = typology_colors) +
  facet_wrap(~ Domain, scales = "free_y") +
  labs(x = "Spherical harmonic degree", y = "Mean power (proportion)") +
  theme_bw() +
  theme(panel.grid = element_blank())
```

## 2. Do reduction strategies differ?

PCA is descriptive. To test whether the strategies differ in their SH descriptors we use **PERMANOVA** (Anderson 2001), implemented in `vegan::adonis2()`. PERMANOVA partitions a distance matrix among groups and assesses the group effect by permutation, so it copes gracefully with many variables and few specimens.

```{r permanova}
library(vegan)

grp_df <- data.frame(Strategy = groups)

set.seed(1)
pm_scar  <- adonis2(dist(scar_ilr)  ~ Strategy,
                    data = grp_df, permutations = 999)
pm_morph <- adonis2(dist(morph_ilr) ~ Strategy,
                    data = grp_df, permutations = 999)
```

```{r permanova-table, echo = FALSE}
permanova_summary <- data.frame(
  Domain    = c("Scar direction", "Morphology"),
  R2        = c(pm_scar$R2[1],        pm_morph$R2[1]),
  `Pseudo-F`= c(pm_scar$F[1],         pm_morph$F[1]),
  `p-value` = c(pm_scar$`Pr(>F)`[1],  pm_morph$`Pr(>F)`[1]),
  check.names = FALSE)

knitr::kable(permanova_summary, digits = 3,
  caption = "PERMANOVA for differences among reduction strategies, by domain")
```

The scar direction PERMANOVA returns p = `r if (can_run) round(pm_scar[["Pr(>F)"]][1], 3)`, with R<sup>2</sup> = `r if (can_run) round(pm_scar[["R2"]][1], 3)` of the variation in scar pattern attributable to reduction strategy. In other words, knowing core types tell us a meaningful amount about its scar pattern signature.

### Which reduction strategies differ?

The omnibus PERMANOVA tells us that the strategies differ, but not which pairs. For that we run PERMANOVA on each pair of strategies and correct the p-values for multiple comparisons:

```{r pairwise-permanova}
pairwise_permanova <- function(ilr_df, grp, permutations = 999,
                               p.adjust.method = "holm") {
  grp   <- droplevels(factor(grp))
  combs <- utils::combn(levels(grp), 2)
  res <- lapply(seq_len(ncol(combs)), function(k) {
    keep    <- grp %in% combs[, k]
    sub_grp <- droplevels(grp[keep])
    a <- vegan::adonis2(dist(ilr_df[keep, , drop = FALSE]) ~ sub_grp,
                        data = data.frame(sub_grp), permutations = permutations)
    data.frame(pair = paste(combs[1, k], "vs", combs[2, k]),
               R2 = a$R2[1], F = a$F[1], p = a[["Pr(>F)"]][1])
  })
  out       <- do.call(rbind, res)
  out$p_adj <- p.adjust(out$p, method = p.adjust.method)
  out
}

set.seed(1); pw_scar  <- pairwise_permanova(scar_ilr,  groups)
set.seed(1); pw_morph <- pairwise_permanova(morph_ilr, groups)
```

```{r pairwise-table, echo = FALSE}
pw_all <- rbind(
  cbind(Domain = "Scar direction", pw_scar),
  cbind(Domain = "Morphology",     pw_morph))

knitr::kable(pw_all, digits = 3, row.names = FALSE,
  caption = "Pairwise PERMANOVA between reduction strategies. p_adj: Holm-Bonferroni correction within each domain.")
```

Read `p_adj` rather than the raw `p` to decide which pairs differ: with five strategies there are ten comparisons per domain, and the correction guards against false positives from repeated testing. We use the **Holm-Bonferroni** step-down correction (`method = "holm"`), which controls the family-wise error rate (FWER) while being uniformly more powerful than plain Bonferroni; switch `p.adjust.method` to `"BH"` for a less conservative false-discovery-rate criterion.

## 3. Is one strategy more standardised?

"Difference in location" (PERMANOVA, above) is not the same as "difference in variability". A central question in lithic studies concerns how standardised a reduction strategy is, which is a question about variability: a more standardised strategy produces a tighter cluster of specimens.

`vegan::betadisper()` is the multivariate analogue of Levene's test: it measures each specimen's Aitchison distance to its group centroid and tests whether those distances differ among groups (Anderson 2006). Smaller average distance = tighter cluster = more standardised.

```{r betadisper}
# type = "centroid" gives the original PERMDISP of Anderson (2006): distances are
# measured to the group centroid (betadisper's default is the spatial median).
bd_scar  <- betadisper(dist(scar_ilr),  groups, type = "centroid")
bd_morph <- betadisper(dist(morph_ilr), groups, type = "centroid")

set.seed(1); pt_scar  <- permutest(bd_scar,  permutations = 999)
set.seed(1); pt_morph <- permutest(bd_morph, permutations = 999)
```

```{r betadisper-table, echo = FALSE}
disp_summary <- data.frame(
  Domain    = c("Scar direction", "Morphology"),
  `F`       = c(pt_scar$tab$F[1],         pt_morph$tab$F[1]),
  `p-value` = c(pt_scar$tab$`Pr(>F)`[1],  pt_morph$tab$`Pr(>F)`[1]),
  check.names = FALSE)

knitr::kable(disp_summary, digits = 3,
  caption = "Homogeneity of multivariate dispersion (PERMDISP) by domain")
```

```{r betadisper-plot, fig.cap = "Distance to group centroid by reduction strategy and domain. Lower and tighter = more standardised."}
disp_df <- rbind(
  data.frame(Domain = "Scar direction", Strategy = groups, distance = bd_scar$distances),
  data.frame(Domain = "Morphology",     Strategy = groups, distance = bd_morph$distances))
disp_df$Domain   <- factor(disp_df$Domain, levels = c("Scar direction", "Morphology"))
disp_df$Strategy <- factor(disp_df$Strategy, levels = strategy_levels)

ggplot(disp_df, aes(Strategy, distance, colour = Strategy, fill = Strategy)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.25, linewidth = 0.35) +
  geom_jitter(width = 0.15, size = 1.6, alpha = 0.7, shape = 16) +
  stat_summary(fun = mean, geom = "point", shape = 16, size = 2.4, colour = "white") +
  scale_colour_manual(values = typology_colors) +
  scale_fill_manual(values  = typology_colors) +
  facet_wrap(~ Domain, scales = "free_y") +
  labs(x = NULL, y = "Distance to group centroid") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 20, hjust = 1),
        legend.position = "none")
```

A significant `permutest` p-value (table above) means the strategies are not equally variable in that domain; the boxplots show which are tighter (more standardised) and which are more dispersed. For a direct ranking within a domain, average each group's distance to its centroid, e.g. `tapply(bd_scar$distances, groups, mean)` for scar direction.

**Read the two tests together.** PERMANOVA and PERMDISP are not independent: a PERMANOVA is sensitive to differences in dispersion as well as in location, so when `betadisper` is also significant, a significant PERMANOVA can no longer be read as unambiguous evidence of a location difference. The groups may differ in spread, in centroid, or both. In that situation interpret the group "difference" cautiously and lean on the ordination and the dispersion boxplots to tell the two apart (Anderson & Walsh 2013).

## Reproducibility

PERMANOVA and `permutest` are randomised, so call `set.seed()` immediately before each to get reproducible output (as we did above). Increasing `permutations` tightens the p-value estimate at the cost of run time.

## Where to go next

- Swap in the full coefficient vectors (`include_coeffs = TRUE` in `spharm_to_dataframe()`) for a finer-grained, alignment-dependent analysis. These are real value (not compositional), so use them directly with Euclidean distances.
- Add specimen metadata (raw material, layer, reduction intensity…) and test those factors the same way.
- Use `export_spharm_html()` to inspect, in 3D, the specimens that sit at the extremes of a PC axis or a dispersion boxplot.

## References

Aitchison, J. (1986). The Statistical Analysis of Compositional Data. London: Chapman & Hall.

Anderson, M. J. (2001). A new method for non-parametric multivariate analysis of variance. Austral Ecology, 26(1), 32–46.

Anderson, M. J. (2006). Distance-based tests for homogeneity of multivariate dispersions. Biometrics, 62(1), 245–253.

Anderson, M. J., & Walsh, D. C. I. (2013). PERMANOVA, ANOSIM, and the Mantel test in the face of heterogeneous dispersions: What null hypothesis are you testing? Ecological Monographs, 83(4), 557–574.

Egozcue, J. J., Pawlowsky-Glahn, V., Mateu-Figueras, G., & Barceló-Vidal, C. (2003). Isometric logratio transformations for compositional data analysis. Mathematical Geology, 35(3), 279–300.

Wieczorek, M. A., & Meschede, M. (2018). SHTools: Tools for working with spherical harmonics. Geochemistry, Geophysics, Geosystems, 19(8), 2574–2592.

Xiao, P., Li, H., & Marwick, B. (in preparation). Characterizing core morphology and scar patterning within a unified spherical harmonic framework.
