---
title: "Statistical analysis of spherical harmonic data"
author: "Peiyuan Xiao, Ben Marwick"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Statistical analysis of spherical harmonic data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# This vignette uses several Suggested packages. If any are missing we still
# render the prose and show the code, but skip evaluation, so the vignette
# never fails to build.
can_run <- requireNamespace("vegan",        quietly = TRUE) &&
           requireNamespace("compositions", quietly = TRUE) &&
           requireNamespace("ggplot2",      quietly = TRUE) &&
           requireNamespace("tidyr",        quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 6,
  fig.height = 4,
  eval     = can_run
)
```

## Introduction

The companion vignette `vignette("Introduction", package = "spharmlithic")` shows how to turn 3D artefacts into spherical harmonic (SH) coefficients with `spharm_from_directions()` and `spharm_from_meshes()`, and how to flatten the result with `spharm_to_dataframe()`. **This vignette picks up where that one leaves off: you have a table of SH descriptors, one row per specimen — now what?**

We work through three questions that recur in lithic analysis:

1.  **Exploration** — how is shape/scar variation structured across specimens? (Principal Component Analysis.)
2.  **Group difference** — do specimens made by different reduction strategies differ in their SH descriptors? (PERMANOVA.)
3.  **Standardisation** — is one strategy more *standardised* (less variable) than another? (Multivariate dispersion.)

Three features of SH power spectra shape the choice of method:

- **The power spectrum is rotation-invariant.** Each `power_lN` value summarises the energy at spherical-harmonic degree `N`, independent of how the specimen was oriented. This makes the power spectrum a robust feature set for cross-specimen comparison — we use it throughout. (The raw coefficients carry more detail but depend on alignment.)
- **The power spectrum is compositional.** Each specimen's spectrum is *closed* to a constant sum (here, 1): the values are relative proportions of energy across degrees, not absolute magnitudes. Compositional data live on the simplex, not in ordinary Euclidean space, so applying Euclidean-based methods (PCA, distances) to the raw proportions induces spurious negative correlations (the *closure problem*). The fix is a log-ratio transform — we use the **isometric log-ratio (ILR)** transform (Egozcue et al. 2003), which maps the composition into ordinary Euclidean coordinates where standard methods are valid. Distances computed on ILR coordinates are *Aitchison distances*.
- **`p` is often comparable to or larger than `n`.** Even the power spectrum has 20+ columns. Classical MANOVA breaks down when variables approach the number of specimens, so we lean on distance- and permutation-based methods (PERMANOVA, `betadisper`) and dimension reduction (PCA).

## The example data

The package bundles a worked example: spherical harmonic power spectra for **58 experimental cores** knapped with five known reduction strategies, from the authors' experimental reference collection (Xiao, Li, & Marwick, in preparation). Two parallel tables describe the same specimens in two domains — **scar direction** (how flaking removals are organised) and **morphology** (overall shape). Each was produced exactly as in the Introduction vignette (`spharm_from_*()` → `spharm_to_dataframe()`); we ship the flattened tables so this vignette needs no Python backend.

```{r load-data}
scar  <- read.csv(
  system.file("extdata", "exp_cores_scar.csv",  package = "spharmlithic"),
  check.names = FALSE)
morph <- read.csv(
  system.file("extdata", "exp_cores_morph.csv", package = "spharmlithic"),
  check.names = FALSE)

# Put the strategies in a sensible order for plots and tables
strategy_levels <- c("Unidirectional", "Bidirectional",
                     "Levallois", "Discoid", "Multiplatform")
scar$Typology  <- factor(scar$Typology,  levels = strategy_levels)
morph$Typology <- factor(morph$Typology, levels = strategy_levels)

# Shared colour palette, so every figure in this vignette matches the look of
# the source-study figures (Xiao, Li & Marwick, in preparation).
typology_colors <- c(
  "Unidirectional" = "#BA8530",
  "Bidirectional"  = "#788C4A",
  "Levallois"      = "#4A6E8A",
  "Discoid"        = "#802520",
  "Multiplatform"  = "#8A7A68"
)

knitr::kable(
  as.data.frame(table(Strategy = scar$Typology)),
  col.names = c("Reduction strategy", "n"),
  caption   = "Experimental cores per reduction strategy")
```

Each table has an `ID`, a `Typology` (the reduction strategy), and the power spectrum from degree 1 to 20 (`power_l1` … `power_l20`; degree 0, the constant term, is dropped because it carries no shape information). Each row sums to 1:

```{r preview}
knitr::kable(
  head(scar[, 1:6]),
  caption = "First few rows of the scar-direction power spectra")

# Confirm the spectra are closed (compositional)
range(rowSums(scar[, grep("^power_l", names(scar))]))
```

It is convenient to separate the numeric feature matrix from the grouping variable:

```{r feature-matrices}
# Degree selection (matching the source study, Xiao, Li & Marwick in prep.):
# the informative shape/scar structure concentrates in the low spherical-harmonic
# degrees, so we keep degrees 1-6 for scar direction and 1-8 for morphology and
# drop the higher, near-noise degrees before the compositional analysis.
scar_cols   <- paste0("power_l", 1:6)
morph_cols  <- paste0("power_l", 1:8)

scar_feat   <- scar[,  scar_cols]
morph_feat  <- morph[, morph_cols]
groups      <- scar$Typology      # identical IDs/order in both tables
```

## Treating the power spectrum as compositional data

Before any Euclidean-based analysis we move the spectra off the simplex with the ILR transform. ILR (like any log-ratio) requires strictly positive values, but the scar spectra contain exact zeros (degrees with no energy). We handle them with a **multiplicative zero replacement** that preserves the unit sum, following Xiao, Li & Marwick (in preparation):

```{r ilr}
# Replace zeros with a small fraction of the smallest non-zero value in the
# row, then rescale the non-zeros so the row still sums to 1.
replace_zeros <- function(X, delta = NULL) {
  X <- as.matrix(X)
  for (i in seq_len(nrow(X))) {
    row_i    <- X[i, ]
    zero_idx <- row_i == 0
    if (!any(zero_idx)) next
    d  <- if (is.null(delta)) min(row_i[!zero_idx]) * 0.65 else delta
    nz <- sum(zero_idx)
    row_i[zero_idx]  <- d
    row_i[!zero_idx] <- row_i[!zero_idx] * (1 - nz * d)
    X[i, ] <- row_i
  }
  X
}

# Wrap the transform in a helper that first drops any zero-variance degree (a
# column identical across all specimens carries no information and would break
# the log-ratio), then replaces zeros and applies ILR. This mirrors the
# make_ilr() helper in the source study.
make_ilr <- function(power_df) {
  X    <- as.matrix(power_df)
  keep <- apply(X, 2, function(v) sd(v, na.rm = TRUE) > 0)
  as.data.frame(compositions::ilr(replace_zeros(X[, keep, drop = FALSE])))
}

# ILR coordinates: D parts -> (D - 1) real-valued coordinates per specimen.
scar_ilr  <- make_ilr(scar_feat)    # 6 degrees -> 5 ILR coordinates
morph_ilr <- make_ilr(morph_feat)   # 8 degrees -> 7 ILR coordinates

dim(scar_ilr)   # 58 specimens x 5 ILR coordinates
```

All of the analyses below operate on these ILR coordinates. Note that this is specific to the *power spectrum* (a composition); the **raw coefficients** returned by `spharm_to_dataframe(include_coeffs = TRUE)` are ordinary real-valued numbers and would be analysed directly (e.g. with Euclidean distances), not log-ratio transformed.

## 1. Exploring variation with PCA

A quick way to see whether the strategies separate at all is to project each domain's ILR coordinates onto their first two principal components. We run PCA on the **unscaled** coordinates (no `scale. = TRUE`): ILR already lays the composition out in a Euclidean space whose distances are the Aitchison distances the PERMANOVA below relies on, so standardising each coordinate would reweight the parts and distort that geometry.

```{r pca}
pca_scar  <- prcomp(scar_ilr)
pca_morph <- prcomp(morph_ilr)

# Variance explained by the first few components, both domains
var_tab <- data.frame(
  PC = paste0("PC", 1:5),
  `Scar direction (%)` = round(100 * summary(pca_scar)$importance[2, 1:5], 1),
  `Morphology (%)`     = round(100 * summary(pca_morph)$importance[2, 1:5], 1),
  check.names = FALSE)
knitr::kable(var_tab, row.names = FALSE,
  caption = "Variance explained by the leading principal components, by domain")
```

```{r pca-plot, fig.cap = "Cores in the space of the first two principal components of the ILR coordinates, by domain, coloured by reduction strategy, with convex hulls outlining each group. Axis scales are free; variance explained is in the table above."}
library(ggplot2)

pca_df <- rbind(
  data.frame(Domain = "Scar direction",
             PC1 = pca_scar$x[, 1],  PC2 = pca_scar$x[, 2],  Strategy = groups),
  data.frame(Domain = "Morphology",
             PC1 = pca_morph$x[, 1], PC2 = pca_morph$x[, 2], Strategy = groups))
pca_df$Domain   <- factor(pca_df$Domain, levels = c("Scar direction", "Morphology"))
pca_df$Strategy <- factor(pca_df$Strategy, levels = strategy_levels)

# Convex hulls per strategy, matching the project's LDA / ternary scatterplots
hull_df <- do.call(rbind, lapply(
  split(pca_df, list(pca_df$Domain, pca_df$Strategy), drop = TRUE),
  function(d) d[chull(d$PC1, d$PC2), ]))

ggplot(pca_df, aes(PC1, PC2, colour = Strategy)) +
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.35, linetype = "dashed") +
  geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.35, linetype = "dashed") +
  geom_polygon(data = hull_df, aes(fill = Strategy, group = Strategy),
               alpha = 0.25, colour = NA) +
  geom_point(size = 2, alpha = 0.85, shape = 16) +
  scale_colour_manual(values = typology_colors) +
  scale_fill_manual(values  = typology_colors) +
  facet_wrap(~ Domain, scales = "free") +
  labs(x = "PC1", y = "PC2") +
  theme_bw() +
  theme(panel.grid = element_blank())
```

The convex hulls give a first impression of both *location* (do the groups sit in different places?) and *spread* (are some groups tighter than others?) — the two things we test formally below.

It is also informative to look at the **mean power spectrum** of each strategy (on the original closed scale): the degree at which energy concentrates reflects the angular scale of the dominant structure.

```{r spectrum-plot, fig.cap = "Mean power spectrum by reduction strategy and domain (ribbon = +/- 1 SE). Values are closed proportions."}
library(tidyr)

# The two domains now keep different numbers of degrees, so pivot each to long
# form (a shared schema) before combining.
spec_long <- rbind(
  data.frame(Domain = "Scar direction", Strategy = groups, scar_feat,  check.names = FALSE) |>
    tidyr::pivot_longer(cols = all_of(scar_cols),  names_to = "degree", values_to = "power"),
  data.frame(Domain = "Morphology",     Strategy = groups, morph_feat, check.names = FALSE) |>
    tidyr::pivot_longer(cols = all_of(morph_cols), names_to = "degree", values_to = "power"))
spec_long$degree   <- as.integer(sub("power_l", "", spec_long$degree))
spec_long$Domain   <- factor(spec_long$Domain, levels = c("Scar direction", "Morphology"))
spec_long$Strategy <- factor(spec_long$Strategy, levels = strategy_levels)

spec_summ <- aggregate(power ~ Domain + Strategy + degree, spec_long,
                       FUN = function(x) c(m = mean(x),
                                           se = sd(x) / sqrt(length(x))))
spec_summ <- do.call(data.frame, spec_summ)

ggplot(spec_summ, aes(degree, power.m, colour = Strategy, fill = Strategy)) +
  geom_ribbon(aes(ymin = power.m - power.se, ymax = power.m + power.se),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.6) +
  scale_colour_manual(values = typology_colors) +
  scale_fill_manual(values  = typology_colors) +
  facet_wrap(~ Domain, scales = "free_y") +
  labs(x = "Spherical harmonic degree", y = "Mean power (proportion)") +
  theme_bw() +
  theme(panel.grid = element_blank())
```

## 2. Do reduction strategies differ? (PERMANOVA)

PCA is descriptive. To *test* whether the strategies differ in their SH descriptors we use **PERMANOVA** (Anderson 2001), implemented in `vegan::adonis2()`. PERMANOVA partitions a distance matrix among groups and assesses the group effect by permutation, so it copes gracefully with many variables and few specimens.

Dissimilarity is the **Aitchison distance** — ordinary Euclidean distance on the ILR coordinates:

```{r permanova}
library(vegan)

grp_df <- data.frame(Strategy = groups)

set.seed(1)
pm_scar  <- adonis2(dist(scar_ilr)  ~ Strategy,
                    data = grp_df, permutations = 999)
pm_morph <- adonis2(dist(morph_ilr) ~ Strategy,
                    data = grp_df, permutations = 999)
```

```{r permanova-table, echo = FALSE}
permanova_summary <- data.frame(
  Domain    = c("Scar direction", "Morphology"),
  R2        = c(pm_scar$R2[1],        pm_morph$R2[1]),
  `Pseudo-F`= c(pm_scar$F[1],         pm_morph$F[1]),
  `p-value` = c(pm_scar$`Pr(>F)`[1],  pm_morph$`Pr(>F)`[1]),
  check.names = FALSE)

knitr::kable(permanova_summary, digits = 3,
  caption = "PERMANOVA for differences among reduction strategies, by domain")
```

The scar-direction PERMANOVA returns p = `r if (can_run) round(pm_scar[["Pr(>F)"]][1], 3)`, with R<sup>2</sup> = `r if (can_run) round(pm_scar[["R2"]][1], 3)` of the variation in scar-pattern organisation attributable to reduction strategy. In other words, **knowing how a core was reduced tells us a meaningful amount about its scar-direction signature.** Read the morphology row of the table the same way.

### Which strategies differ? (pairwise PERMANOVA)

The omnibus PERMANOVA tells us *that* the strategies differ, but not *which* pairs. For that we run PERMANOVA on each pair of strategies and correct the p-values for multiple comparisons. No extra package is needed — we loop over `vegan::adonis2()` and adjust with `stats::p.adjust()`:

```{r pairwise-permanova}
pairwise_permanova <- function(ilr_df, grp, permutations = 999,
                               p.adjust.method = "holm") {
  grp   <- droplevels(factor(grp))
  combs <- utils::combn(levels(grp), 2)
  res <- lapply(seq_len(ncol(combs)), function(k) {
    keep    <- grp %in% combs[, k]
    sub_grp <- droplevels(grp[keep])
    a <- vegan::adonis2(dist(ilr_df[keep, , drop = FALSE]) ~ sub_grp,
                        data = data.frame(sub_grp), permutations = permutations)
    data.frame(pair = paste(combs[1, k], "vs", combs[2, k]),
               R2 = a$R2[1], F = a$F[1], p = a[["Pr(>F)"]][1])
  })
  out       <- do.call(rbind, res)
  out$p_adj <- p.adjust(out$p, method = p.adjust.method)
  out
}

set.seed(1); pw_scar  <- pairwise_permanova(scar_ilr,  groups)
set.seed(1); pw_morph <- pairwise_permanova(morph_ilr, groups)
```

```{r pairwise-table, echo = FALSE}
pw_all <- rbind(
  cbind(Domain = "Scar direction", pw_scar),
  cbind(Domain = "Morphology",     pw_morph))

knitr::kable(pw_all, digits = 3, row.names = FALSE,
  caption = "Pairwise PERMANOVA between reduction strategies. p_adj: Holm-Bonferroni correction within each domain.")
```

Read `p_adj` rather than the raw `p` to decide which pairs differ: with five strategies there are ten comparisons per domain, and the correction guards against false positives from repeated testing. We use the **Holm-Bonferroni** step-down correction (`method = "holm"`), which controls the family-wise error rate while being uniformly more powerful than plain Bonferroni; switch `p.adjust.method` to `"BH"` for a less conservative false-discovery-rate criterion.

## 3. Is one strategy more standardised? (multivariate dispersion)

"Difference in *location*" (PERMANOVA, above) is not the same as "difference in *variability*". A central question in lithic studies — how *standardised* a reduction strategy is — is a question about variability: a more standardised strategy produces a tighter cluster of specimens.

`vegan::betadisper()` is the multivariate analogue of Levene's test: it measures each specimen's Aitchison distance to its group centroid and tests whether those distances differ among groups (Anderson 2006). **Smaller average distance = tighter cluster = more standardised.** (This is the PERMDISP test reported in the source study.)

```{r betadisper}
# type = "centroid" gives the original PERMDISP of Anderson (2006): distances are
# measured to the group centroid (betadisper's default is the spatial median).
bd_scar  <- betadisper(dist(scar_ilr),  groups, type = "centroid")
bd_morph <- betadisper(dist(morph_ilr), groups, type = "centroid")

set.seed(1); pt_scar  <- permutest(bd_scar,  permutations = 999)
set.seed(1); pt_morph <- permutest(bd_morph, permutations = 999)
```

```{r betadisper-table, echo = FALSE}
disp_summary <- data.frame(
  Domain    = c("Scar direction", "Morphology"),
  `F`       = c(pt_scar$tab$F[1],         pt_morph$tab$F[1]),
  `p-value` = c(pt_scar$tab$`Pr(>F)`[1],  pt_morph$tab$`Pr(>F)`[1]),
  check.names = FALSE)

knitr::kable(disp_summary, digits = 3,
  caption = "Homogeneity of multivariate dispersion (PERMDISP) by domain")
```

```{r betadisper-plot, fig.cap = "Distance to group centroid by reduction strategy and domain. Lower and tighter = more standardised."}
disp_df <- rbind(
  data.frame(Domain = "Scar direction", Strategy = groups, distance = bd_scar$distances),
  data.frame(Domain = "Morphology",     Strategy = groups, distance = bd_morph$distances))
disp_df$Domain   <- factor(disp_df$Domain, levels = c("Scar direction", "Morphology"))
disp_df$Strategy <- factor(disp_df$Strategy, levels = strategy_levels)

ggplot(disp_df, aes(Strategy, distance, colour = Strategy, fill = Strategy)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.25, linewidth = 0.35) +
  geom_jitter(width = 0.15, size = 1.6, alpha = 0.7, shape = 16) +
  stat_summary(fun = mean, geom = "point", shape = 16, size = 2.4, colour = "white") +
  scale_colour_manual(values = typology_colors) +
  scale_fill_manual(values  = typology_colors) +
  facet_wrap(~ Domain, scales = "free_y") +
  labs(x = NULL, y = "Distance to group centroid") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 20, hjust = 1),
        legend.position = "none")
```

A significant `permutest` p-value (table above) means the strategies are *not* equally variable in that domain; the boxplots show which are tighter (more standardised) and which are more dispersed. For a direct ranking within a domain, average each group's distance to its centroid — e.g. `tapply(bd_scar$distances, groups, mean)` for scar direction.

**Read the two tests together.** PERMANOVA and PERMDISP are not independent: a PERMANOVA is sensitive to differences in *dispersion* as well as in *location*, so when `betadisper` is *also* significant, a significant PERMANOVA can no longer be read as unambiguous evidence of a *location* (mean-shape) difference — the groups may differ in spread, in centroid, or both. In that situation interpret the group "difference" cautiously and lean on the ordination and the dispersion boxplots to tell the two apart (Anderson & Walsh 2013).

> **Why not a coefficient-of-variation test here?** Because the power spectrum is closed (every specimen sums to 1), it carries no absolute-size information — the total is constant — and the ILR coordinates are interval-scaled (they can be negative). The coefficient of variation is only meaningful for a ratio-scale, non-negative variable with a true zero, so it applies to neither. To compare the *variability of size* across strategies you would bring in an external measurement (e.g. a linear core dimension or mass) and test it with, for example, the `cvequality` package.

## Reproducibility

PERMANOVA and `permutest` are randomised, so call `set.seed()` immediately before each to get reproducible output (as we did above). Increasing `permutations` tightens the p-value estimate at the cost of run time.

## Where to go next

- Swap in the **full coefficient vectors** (`include_coeffs = TRUE` in `spharm_to_dataframe()`) for a finer-grained, alignment-dependent analysis. These are real-valued (not compositional), so use them directly with Euclidean distances — no ILR.
- Add specimen metadata (raw material, archaeological layer, …) and test those factors the same way — this is how the source study analyses its archaeological core assemblage (Xiao, Li, & Marwick, in preparation).
- Use `export_spharm_html()` to inspect, in 3D, the specimens that sit at the extremes of a PC axis or a dispersion boxplot.

## References

Aitchison, J. (1986). *The Statistical Analysis of Compositional Data*. London: Chapman & Hall.

Anderson, M. J. (2001). A new method for non-parametric multivariate analysis of variance. *Austral Ecology*, 26(1), 32–46.

Anderson, M. J. (2006). Distance-based tests for homogeneity of multivariate dispersions. *Biometrics*, 62(1), 245–253.

Anderson, M. J., & Walsh, D. C. I. (2013). PERMANOVA, ANOSIM, and the Mantel test in the face of heterogeneous dispersions: What null hypothesis are you testing? *Ecological Monographs*, 83(4), 557–574.

Egozcue, J. J., Pawlowsky-Glahn, V., Mateu-Figueras, G., & Barceló-Vidal, C. (2003). Isometric logratio transformations for compositional data analysis. *Mathematical Geology*, 35(3), 279–300.

Wieczorek, M. A., & Meschede, M. (2018). SHTools: Tools for working with spherical harmonics. *Geochemistry, Geophysics, Geosystems*, 19(8), 2574–2592.

Xiao, P., Li, H., & Marwick, B. (in preparation). Characterizing core morphology and scar patterning within a unified spherical harmonic framework.
